---
title: "BSC | Assignment"
author: "Edgar Chacón (edgarpedro.chacon01@estudiant.upf.edu)"
date: "Last update: `r format(Sys.time(), '%d %B, %Y')`"      
output:
  html_document:
    toc: true
    toc_float: true
    fig_caption: true
---

<style>
#TOC .list-group-item.active, .list-group-item.active:focus {
    color: #ffffff !important;
    background-color: #AABF55;
    transition: background-color 0.5s;
}
#TOC .list-group-item.active:hover {
    color: #ffffff !important;
    background-color: #BF9055;
}
#TOC .list-group-item.active .accent {
    color: inherit !important;
}

.title {
  color: #AABF55 !important;
}

h1 {
  color: #AABF55;
  border-bottom: 1px solid #BF9055;
}

h2 {
  color: #BF9055;
}

h3 {
  color: #BF9055;
}

h4 {
  color: #BF9055;
}

h4.date {
  color: black;
  font-weight: normal;
}

h5 {
  color: black;
}

a {
  color: #AABF55; 
  transition: color 0.5s;  
}

a:hover {
  color: #BF9055; 
}

a:visited {
  color: #AABF55;
}

a:visited:hover {
  color: #BF9055;
}

.orange_accent {
  color: #BF9055; 
}

.green_accent {
  color: #AABF55; 
}

.question {
  color: black;
  font-size: 14px;
  font-style: italic;
  font-weight: bold;
}

pre {
  background-color: #E9ECDD;
  padding: 10px;
}

.caption-title {
  color: #AABF55 !important;
}

.table_title {
  color: #BF9055 !important;
}

.horizontal-scroll {
  margin-right: 150px !important;
  margin-left: 65px !important;
  padding-top: 5px !important;
}

.table {
  max-width: 630px !important;
  padding-left: 20px !important;
  padding-right: 20px !important;
  padding-top: 20px !important;
  padding-bottom: 20px !important;
  margin-right: 0px !important;
  margin-left: 0px !important;
  border-top: 2px solid #ECE6DD !important;
  border-bottom: 2px solid #ECE6DD !important;
}

caption {
  padding-left: 5px !important;
  padding-right: 0px !important;
  border-left: 2px solid #ECE6DD !important;
  margin-left: 0px !important;
}

.figure {
  margin-right: 150px !important;
  margin-left: 0px !important;
}

.img.widefigure{
  padding-left: 5px !important;
  padding-right: 155px !important;
  padding-top: 5px !important;
  padding-bottom: 5px !important;
  margin-right: 150px !important;
  margin-left: 0px !important;
}

img {
  padding-left: 10px !important;
  padding-right: 10px !important;
  padding-top: 10px !important;
  padding-bottom: 10px !important;
  border: 2px solid #E9ECDD !important;
  border-radius: 4px !important;
  margin-right: 150px !important;
  margin-left: 0px !important;
}

p.caption {
  margin-left: 65px !important;
  padding-left: 5px !important;
  padding-right: 0px !important;
  border-left: 2px solid #E9ECDD !important;
}

</style>

```{r SETUP, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<br>

# <span class="accent">0.</span> Introduction

This file corresponds to the assignment asked for the application to the "<span class="orange_accent">Computational Biology Group</span>" at the <span class="orange_accent">Barcelona Supercomputing Center</span> (<b class="green_accent">BSC</b>).
<br>
<br>
This assignment consist of the following tasks:
<br>

  1. Download The PDB files
  <br>
  2. Install frustratometeR
  <br>
  3. Calculate single residue frustration for all PDBs
  <br>
  4. Calculate a histogram of frustration per residue for all PDBs
  <br>
  5. Calculate the frequency of amino acids in all PDBs

<br>

<span class="orange_accent"><b>NOTE</b>: Absolute paths in this R  markdown may not work properly for others; adjust if necessary.</span>

## <span class="accent">0.1.</span> Libraries

Libraries used in this R markdown.
```{r LIBRARIES, message=FALSE, warning=FALSE}
# ---------------------------------------------------------------------LIBRARIES

invisible(lapply(c("usethis", "devtools", "reticulate", "ggplot2", "dplyr"), library, character.only = TRUE))

devtools::install_github("proteinphysiologylab/frustratometeR")

library(frustratometeR)
```

Modeller had to be installed previously.
<br>
Specific information about dependencies installed in order to run <span class="orange_accent">`frustratometeR`</span> are in the ["Install frustratometeR" section](#section2). 

<!--
The following code chunk is hidden because its lines are not that relevant
to the final markdown. Their purpose is to set some colors that can be used for plots.


## <span class="accent">0.2.</span> Colors

Color variables used in this R markdown.
--->

```{r COLORS, include=FALSE}
# ------------------------------------------------------------------------COLORS

my_orange <- c("#BF9055")
my_green <- c("#90A430")
```

<br>

<a name="section1"></a>

# <span class="accent">1.</span> Download The PDB files

To download the corresponding PDB files associated with the PDB IDs listed in the file [List_100_PDBs](./List_100_PDBs), first I had to extract those IDs from the list.

```{r, message=FALSE, results='hide'}
# -------------------------------------------------------------PDB ID EXTRACTION

pdb_list <- read.table("./List_100_PDBs", header = TRUE)
pdb_ids <- pdb_list$PdbID
pdb_ids

```

Then, generate the download URLs.

```{r, message=FALSE, results='hide'}
# --------------------------------------------------------GENERATE DOWNLOAD URLS

base_url <- "https://files.rcsb.org/download/"
download_urls <- paste0(base_url, pdb_ids, ".pdb")
download_urls

```

And finally, use the <span class="orange_accent">`wget`</span> command-line utility to download all the listed PDB files (`r length(pdb_ids)`).

```{r}
# --------------------------------------------------------DOWNLOAD THE PDB FILES

pdb_folder <- "./PDBs"
if (!file.exists(pdb_folder)) {
  dir.create(pdb_folder, showWarnings = FALSE)
}

for (pdb_id in pdb_ids) {
  wget_command <- paste("wget", paste0("https://files.rcsb.org/download/", pdb_id, ".pdb"), 
                        "-P", pdb_folder, "-nc --progress=bar:force:noscroll -q --show-progress")
  system(wget_command)
}

num_files <- system(paste("find", pdb_folder, "-name '*.pdb' | wc -l"), intern = TRUE)
cat(paste(num_files, "PDB files have been downloaded successfully.\n"))

```

<br>

<a name="section2"></a>

# <span class="accent">2.</span> Install frustratometeR

To install <span class="orange_accent">`frustratometeR`</span>, first I had to install all the necessary dependencies (System, Python and R Package dependencies). 
<br>
To do so, I first executed in terminal the script [install_frustraR.sh](./install_frustraR.sh) obtained from the [GitHub page](https://github.com/proteinphysiologylab/frustratometeR.git) of the R packacge <span class="orange_accent">`frustratometeR`</span>.
<br>

Then, I created a conda environment with python 3.9, where I installed <span class="orange_accent">`modeller`</span> (<span class="green_accent">modeller_10.5-1_amd64.deb</span>) and set the following configuration for this R markdown:

```{r, warning=FALSE}
# --------------------------------------------------MODELLER CONFIGURATION (ENV)

use_python("/home/user/miniconda3/envs/frustratometeR/bin/python3")

use_condaenv(condaenv = "frustratometeR", conda = "/home/user/miniconda3/condabin/conda")

Sys.setenv(RETICULATE_PYTHON = "/home/user/miniconda3/envs/frustratometeR/bin/python3")

reticulate::py_config()

```

To check if <span class="orange_accent">`modeller`</span> actually works, I used the following python code:

```{python}
# ------------------------------------------CHECK MODELLER INSTALLATION {python}

try:
    from modeller import *
    print("Modeller is installed and can be imported successfully.")
except ImportError as e:
    print("ImportError:", e)

```

This allowed me to run the <span class="orange_accent">`frustratometeR`</span> code on the next section ["Calculate single residue frustration for all PDBs"](#section3) and obtain the desired results.

<br>

<a name="section3"></a>

# <span class="accent">3.</span> Calculate single residue frustration for all PDBs

To calculate the single residue frustrations for each PDB file, first I created a <span class="green_accent">`RESULTS folder`</span> where the output files will be stored. 
<br>
Also, I checked which chains were of my interest for the analysis looking at the chains specified on the file [List_100_PDBs](./List_100_PDBs).

```{r}
# ---------------------------------------------RESULTS FOLDER & CHAIN DEFINITION

results_folder <- "/home/rstudio/myfiles/BSC_Assignment/BSC_Assignment/RESULTS/"
if (!file.exists(results_folder)) {
  dir.create(results_folder, showWarnings = FALSE)
}

# NOTE: Adjust the absolute path if necessary.

pdb_chain <- unique(pdb_list$Chain)

```

All chains of interest in the [PDB list](./List_100_PDBs) resulted in the same one: "`r pdb_chain`".
<br>
So, I set that value (<span class="orange_accent">`pdb_chain`</span>) as the desired chain on the following code that will calculate all single residue frustrations for all PDBs found in the <span class="green_accent">`PDBs folder`</span>.

```{r, message=FALSE}
# --------------------------------------------------------CALCULATE FRUSTRATIONS

dir_frustration(PdbsDir = "/home/rstudio/myfiles/BSC_Assignment/BSC_Assignment/PDBs/",
                OrderList = NULL,
                Chain = pdb_chain,
                Mode = "singleresidue",
                ResultsDir = results_folder)

# NOTE: Adjust the absolute path if necessary.

```

All frustration values were successfully generated and stored on the <span class="green_accent">`RESULTS folder`</span>.

<br>

<a name="section4"></a>

# <span class="accent">4.</span> Calculate a histogram of frustration per residue for all PDBs

To generate a histogram showing frustration per residue for all PDBs, I had to take into account that because the analysis was derived from `r num_files` PDB files, I had to firstly combine the data. 
<br>

To do so, I loaded the data obtained when I calculated the frustration of all PDBs, and combined all the resulting data into a single data frame.

```{r}
# --------------------------------------------------------------FRUSTRATION DATA

data_list <- list()

for (pdb_id in pdb_ids) {
  
  file_path <- file.path(results_folder, paste0(pdb_id, "_", pdb_chain, ".done/FrustrationData/", pdb_id, "_", pdb_chain, ".pdb_singleresidue"))
  
  if (file.exists(file_path)) {
    data <- read.table(file_path, header = TRUE, stringsAsFactors = FALSE)
    
    data_list[[length(data_list) + 1]] <- data
  } else {
    message(paste("File not found:", file_path))
  }
}

combined_data <- bind_rows(data_list)

```

And then I generated the histogram itself.

```{r}    
# ---------------------------------------------------FRUSTRATION INDEX HISTOGRAM

ggplot(combined_data, aes(x = FrstIndex)) +
  geom_histogram(binwidth = 0.2, fill = my_green, alpha = 0.6, color = alpha("white", 0.5)) +
  labs(title = "Histogram of Frustration Index for all PDB files",
       x = "Frustration Index",
       y = "Frequency") +
  scale_x_continuous(breaks = seq(floor(min(combined_data$FrstIndex)), ceiling(max(combined_data$FrstIndex)), by = 0.5)) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

```

Because I wasn't sure that the previous plot was the histogram that I had to do, I tried to generate another histogram showing the frustration per residue.
<br>

To do that, I had to aggregate the data calculating the mean frustration score for each residue. 


```{r}
# ---------------------------------------------FRUSTRATION PER RESIDUE HISTOGRAM

frustration_summary <- combined_data %>%
  group_by(AA) %>%
  summarize(avg_frustration = mean(FrstIndex))

frustration_summary$color <- ifelse(frustration_summary$avg_frustration >= 0, my_green, my_orange)

ggplot(frustration_summary, aes(x = AA, y = avg_frustration, fill = color)) +
  geom_bar(stat = "identity", alpha = 0.6) +
  labs(x = "Residue", y = "Average Frustration Index", title = "Frustration per Residue Histogram for all PDB files") +
  scale_fill_identity() +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

```

Finally, as suggested by Gonzalo, I add here the code I previously used to generate a histogram for the frustration results of each one of the PDB files individually.
<br>

First, I created the <span class="green_accent">`PLOTS folder`</span> to store all `r num_files` plots.

```{r}    
# ------------------------------------------------------------------PLOTS FOLDER

plots_folder <- "./PLOTS/"
if (!file.exists(plots_folder)) {
  dir.create(plots_folder, showWarnings = FALSE)
}

```

Then generated and stored each one of the `r num_files` individual frustration index histograms.

```{r}    
# ---------------------------------------INDIVIDUAL FRUSTRATION INDEX HISTOGRAMS

for (pdb_id in pdb_ids) {

  file_path <- file.path(results_folder, paste0(pdb_id, "_", pdb_chain, ".done/FrustrationData/", pdb_id, "_", pdb_chain, ".pdb_singleresidue"))
  
  if (file.exists(file_path)) {
   
    individual_data <- read.table(file_path, header = TRUE, stringsAsFactors = FALSE)
    
    individual_plot <- ggplot(individual_data, aes(x = FrstIndex)) +
      geom_histogram(binwidth = 0.1, fill = my_green, alpha = 0.6, color = alpha("white", 0.5)) +
      labs(title = paste("Histogram of Frustration Index for", pdb_id),
           x = "Frustration Index",
           y = "Frequency") +
      scale_x_continuous(breaks = seq(floor(min(combined_data$FrstIndex)), 
                                      ceiling(max(combined_data$FrstIndex)), by = 0.5)) +
      theme_minimal() +
      theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
    
    output_file <- file.path(plots_folder, paste0(pdb_id, "_", pdb_chain, "_FrstIndex_Histogram.jpg"))
    
    ggsave(output_file, plot = individual_plot, width = 8, height = 6)
    
  } else {
    message(paste("File not found:", file_path))
  }
}

```

All individual frustration index histograms can be accessed in the <span class="green_accent">`PLOTS folder`</span>.

<br>

<a name="section5"></a>

# <span class="accent">5.</span> Calculate the frequency of amino acids in all PDBs

To calculate the frequency of amino acids in all `r num_files` PDB files, I assumed that only the chains specified in the file [List_100_PDBs](./List_100_PDBs) were the ones of interest (Ignoring other possible chains being in some PDBs). 
<br>
Therefore, to save resources, instead of starting from scratch with a count of amino acids extracted from the PDB files (which is too computationally demanding in terms of execution time), I decided to utilize the data resulting from the calculation of frustrations per single residue, which I already had combined.
<br>

Then plotted the histogram of amino acid frequencies.

```{r}
# ------------------------------------------------AMINO ACID FREQUENCY HISTOGRAM

aa_counts <- combined_data %>%
  count(AA) %>%
  arrange(desc(n))

ggplot(aa_counts, aes(x = AA, y = n)) +
  geom_bar(stat = "identity", fill = my_green, alpha = 0.6) +
  labs(x = "Amino Acid", y = "Frequency", title = "Frequency of Each Amino Acid in all PDB files") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

```

<br>

# Session info {.unnumbered}

```{r, results='asis',  echo=FALSE, message=FALSE }
sessionInfo()
```
